---
title: "Why Australia's Housing Boom Outpaced Incomes"
subtitle: "Exploring three decades of imbalance between wages, homes, and capital gains"
author: "Kim Yen Le"
date: "2025-11-05"
format: html
execute:
  echo: true
  warning: false
  message: false
vignette: >
  %\VignetteIndexEntry{Example analysis – housing boom}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(housepack)
library(ggplot2)
library(dplyr)
library(plotly)

theme_housepack <- ggplot2::theme_minimal(base_size = 13) +
    ggplot2::theme(
      plot.title = ggplot2::element_text(
        face = "bold", hjust = 0.5, size = 12
      ),
      plot.subtitle = ggplot2::element_text(
        hjust = 0.5, size = 10, colour = "grey30"
      ),
      axis.title = ggplot2::element_text(
        size = 10, face = "plain", colour = "grey25"
      ),
      axis.text = ggplot2::element_text(
        size = 9, colour = "grey30"
      ),
      panel.background = ggplot2::element_rect(fill = "transparent", colour = NA),
      plot.background  = ggplot2::element_rect(fill = "transparent", colour = NA),
      panel.grid       = ggplot2::element_blank(),
      axis.line        = ggplot2::element_line(colour = "grey50"),
      legend.title     = ggplot2::element_blank(),
      legend.position  = "bottom",
      legend.text      = ggplot2::element_text(size = 9)
    )
```

## Introduction

In 2024, Guardian Australia asked a big question:  
> *"How did Australia's housing market get so bad - and is it all negative gearing's fault?"*

For decades, Australian home prices have climbed far faster than wages. This short analysis uses the `housepack` data and tools to examine how **real disposable income, detached house prices, and capital gains** evolved since 1985. The goal is to visualise the long-term divergence and discuss what it implies for affordability.

## Income vs House Prices

We start by comparing household income and detached-house prices on a base-100 index (1990 = 100).

```{r, message=FALSE, warning=FALSE}
compare_trends(income_pc, detached_full, base_year = 1990)
```

The figure shows that while both series rose after the early-1990s recession, **housing prices surged far ahead of income**, especially through the 2000s and again post-2020. By 2024, the national house-price index is **roughly three times its 1990 level**, whereas income per capita barely quadrupled.

## Capital Gains: Who Benefited?

Housing appreciation also fuels capital gains for owners.

The `cgt_full` dataset tracks **average taxable capital gains** among individuals.

```{r, message=FALSE, warning=FALSE}
ggplot(cgt_full, aes(x = Year, y = Total_gain_adj)) +
geom_col(fill = "#1c4c74") +
labs(
title = "Capital Gains from Property Sales",
y = "AUD ('000)", x = ""
) +
theme_housepack
```

Capital gains spiked in the mid-2000s and again during the pandemic housing surge. This pattern reinforces the **wealth divide** between those who already owned property and those trying to enter the market.

## Regional Divergence

Not every state experienced the boom equally.

We can inspect 2024 values from `detached_ful` to map where housing pressure is greatest.

```{r, message=FALSE, warning=FALSE}
library(sf)
library(rnaturalearth)
aus_map <- rnaturalearth::ne_states(country = "australia", returnclass = "sf") |>
dplyr::filter(!name %in% c("Ashmore and Cartier Islands", "Coral Sea Islands"))

latest_year <- detached_full |> filter(Year == 2024) |>
select(Year, Quarter, Sydney:Canberra) |>
summarise(across(Sydney:Canberra, mean, na.rm = TRUE)) |>
tidyr::pivot_longer(everything(), names_to = "City", values_to = "Index") |>
mutate(State = recode(City,
"Sydney" = "New South Wales", "Melbourne" = "Victoria",
"Brisbane" = "Queensland", "Adelaide" = "South Australia",
"Perth" = "Western Australia", "Hobart" = "Tasmania",
"Darwin" = "Northern Territory", "Canberra" = "Australian Capital Territory"
))

map_data <- aus_map |> left_join(latest_year, by = c("name" = "State"))

ggplot(map_data) +
geom_sf(aes(fill = Index), color = "white", linewidth = 0.3) +
scale_fill_viridis_c(option = "mako", na.value = "grey90") +
labs(title = "Detached-house Price Index by State (2024)", fill = "Index") +
theme_minimal(base_size = 12) +
theme( plot.title = ggplot2::element_text(
        face = "bold", hjust = 0.5, size = 12),
       panel.grid = ggplot2::element_blank())
```

New South Wales leads the pack with an index near 250, far above the 170–190 range seen in most other states. This confirms that the affordability crisis is **geographically concentrated.**

## What Does It All Mean?

- Real wages have grown slowly compared with house prices.

- Taxable capital gains have risen sharply, reflecting windfalls for property investors.

- Regional inequality has deepened - Sydney and Melbourne dominate the boom.

In short, Australia’s housing market has decoupled from incomes. While structural factors like credit, zoning, and population growth play major roles, **tax policy has also amplified the imbalance**.

The rise in taxable capital gains aligns with decades of generous investment incentives. **Negative gearing** allows investors to offset property losses against other income, encouraging leveraged investment in housing. Meanwhile, the **capital-gains tax discount** (introduced in 1999) halves the taxable portion of long-term gains, further rewarding speculation over owner-occupation.

Our data shows the *outcome* - income lagging far behind housing wealth - but these policies help explain *why* housing has become such a lucrative, unequal market.

## Explore It Yourself

Launch the interactive dashboard directly from the package:

```{r, eval = FALSE}
run_housepack()
```

The Shiny app allows users to:

- Toggle between income, housing, and capital-gains views

- Explore yearly and regional changes

- Compare states interactively on a map

## References

Guardian Australia. (2023). *How did Australia's housing market get so bad, and is it all negative gearing's fault?* [Video]. YouTube.  
<https://www.youtube.com/watch?v=iqUFERBJwoY>
