---
title: "Getting Started with HousePack"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{getting-started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
library(housepack)
library(ggplot2)
```

## Overview

`housepack` provides cleaned datasets and visual tools to explore how Australia's housing market, disposable income, and capital gains have evolved from **1985 to 2024**. It includes both static plotting functions and an interactive Shiny dashboard.

## Datasets included

| Dataset | Description |
|----------|-------------|
| `income_pc` | Real household disposable income per capita (quarterly) |
| `detached_full` | Detached house price index for all states and Australia overall |
| `cgt_full` | Average capital gains for taxable individuals (1985–2023) |

## Comparing income and housing prices

The `compare_trends()` function normalises both income and housing indices to a base of 100 (default = 1990), showing how house prices outpaced income growth.

```{r}
compare_trends(income_pc, detached_full)
```

## Launching the interactive Shiny app

The package includes a built-in Shiny dashboard:

```{r, eval = FALSE}
run_housepack()
```

The app allows you to:

- Compare national income, housing, and capital gains

- View state-level differences via a choropleth map

- Analyse long-term trends across regions

## Example interpretation

By 2024, New South Wales reached a detached-house index near 250, while most other states remained between 170 and 190. This reflects how housing affordability pressures are most intense around Sydney.

## Summary

`housepack` connects household income, housing prices, and capital gains in one cohesive toolkit for exploratory analysis of Australia's housing market.

